\section{Introduction to Ordinals}

As children, we are introduced to two notions of counting: the \textit{cardinal numbers} and the \textit{ordinal numbers}. We are taught that the \textit{cardinal numbers} are the numbers of \textit{quantity}---that is, ``one'', ``two'', ``three'', and so on---whereas the \textit{ordinal numbers} are the numbers of \textit{ordering}---that is, ``first'', ``second'', ``third'', and so on. It turns out there is a reason for this. While the cardinal numbers represent different possible \textit{cardinalities} of finite sets, it turns out that the ordinal numbers represent a broader notion of ordering that is not limited to the set of natural numbers.

Broadly speaking, ordinals are nice well-ordered sets that mimic the properties of the natural numbers. The idea is to try and replicate the property in $\N$---at least, when $\N$ is defined using the successor $\dg$ (cf. \Cref{Ch3:Def:succ_dg})---that the ordering comes from both membership and inclusion.

\subsection{Transitive Sets}

We begin by defining the notion of transitive sets.

\begin{boxdefinition}[Transitivity]
    A set $X$ is \textbf{transitive} if every element of $X$ is also a subset of $X$, that is, if
    \begin{align*}
        \parenth{\forall y}\parenth{\parenth{y \subset x} \to \parenth{y \in X}}
    \end{align*}
    holds in $X$.
    % NOTE TO SELF: Include convention for \subset symbol
\end{boxdefinition}

We have examples of transitive sets that come from the set of natural numbers.

\begin{boxexample}\label{Ch3:Eg:3_trans}
    The set $3 = \set{0, 1, 2} = \set{\emptyset, \set{\emptyset}, \set{\emptyset, \set{\emptyset}}}$ is transitive. For example, $2 \in 3$ and $2 \subset 3$ both hold.
\end{boxexample}

However, not all subsets of the naturals are transitive.

\begin{boxnexample}
    $\set{0, 2}$ is not a transitive set. $2 \in \set{0, 2}$ but $2 \not\subset \set{0, 2}$.
\end{boxnexample}

The reason why \Cref{Ch3:Eg:3_trans} works is the following.

\begin{boxlemma}\label{Ch2:Lemma:succ_trans}
    If $X$ is a transitive set, then so is $X\dg = X \cup \set{X}$.
\end{boxlemma}
\begin{proof}
    \sorry
\end{proof}

It does, of course, remain to show that $2$ (and by recursion, $1$ and $0$) are all transitive to show that $3$ is, indeed, transitive, but this is easily done. In fact, we can go one step further.

\begin{boxtheorem}\label{Ch3:Thm:Nat_trans}
    $\N$ is transitive.
\end{boxtheorem}
\begin{proof}
    \sorry % Induction - but use the definition of \N instead of induction as we know it!!!!!!!!
\end{proof}

We are now ready to define ordinals.

\subsection{Ordinals: The Fundamentals}

The idea is for an ordinal to be a transitive set in which the ordering and membership relations are related. This is true in the natural numbers, for example, provided we define the successor function the way we have defined the $\dg$ operation (cf. \Cref{Ch3:Def:succ_dg}).

\begin{boxdefinition}[Ordinal]\label{Ch3:Def:Ordinal}
    A set $\alpha$ is an \textbf{ordinal} if
    \begin{enumerate}
        \item $\alpha$ is a transitive set
        \item The relation $<$ on $\alpha$ given by
        \begin{align*}
            x < y \iff x \in y
        \end{align*}
        for all $x, y \in \alpha$ is a strict well-ordering on $\alpha$.
    \end{enumerate}
\end{boxdefinition}

Here, we pause to make an important observation. In \Cref{Ch3:Def:Ordinal}, we \textbf{explicitly disallow $\alpha \in \alpha$}. The reason for this is that if we had $\alpha \in \alpha$, we would have $\alpha < \alpha$, which contradicts strictness. It is therefore supremely important that the strictness of the well-ordering be included in the definition of an ordinal.

We now relate ordinals and the successor $\dg$.

\begin{boxlemma}
    If $\alpha$ is an ordinal, then $\alpha\dg = \alpha \cup \set{\alpha}$ is also an ordinal.
\end{boxlemma}
\begin{proof}
    \sorry
\end{proof}

We can, in fact, show that $\N$ is an ordinal. However, we are not yet ready to do this. We need more machinery.

We have the following properties of ordinals that follow from the definition.

\begin{boxproposition}[Relationships between Ordinals]
    Let $\alpha$ and $\beta$ be sets.
    \begin{enumerate}
        \item If $\alpha$ is an ordinal and $\beta \in \alpha$, then $\beta$ is an ordinal.
        \item If $\alpha$ and $\beta$ are both ordinals and $\alpha \subsetneq \beta$, then $\alpha \in \beta$.
    \end{enumerate}
\end{boxproposition}
\begin{proof}\hfill
    \begin{enumerate}
        \item \sorry

        \item \sorry
    \end{enumerate}
\end{proof}

It turns out the above properties allow us to compare ordinals and define a notion of ordering between them.

\subsection{Ordering Ordinals}

While there are too many ordinals for there to be a \textit{set} of all ordinals, we can reason collectively about the \textbf{class of all ordinals}, where the word ``class'' is used in the category theoretic sense.  In fact, one can show that this class is well-ordered.

\sorry

\subsection{Transfinite Induction}

In this subsection, we explore a proof strategy that resembles mathematical induction but that works on ordinals instead of the natural numbers. Note that at first glance, it appears that we have ignored the ``base case'' analogue for ordinals, but observe that \eqref{Ch3:Eq:Transfinite_Induction_Condition} includes the case where $\alpha = \emptyset$.

\begin{boxtheorem}[The Principle of Transfinite Induction]\label{Ch3:Thm:Transfinite_Induction}
    Suppose $P(\cdot)$ is a first-order formula in the language $\L$ of sets with a free variable. If, for all ordinals $\alpha$, we have the property that
    \begin{align}
        \text{If } P(\beta) \text{ holds for all ordinals } \beta < \alpha \text{, then } P(\alpha) \text{ holds.}
        \label{Ch3:Eq:Transfinite_Induction_Condition}
        \tag{\ding{37}}
    \end{align}
\end{boxtheorem}
\begin{proof}
    \sorry
\end{proof}

There are many results that can be proven using transfinite induction. Here is one example.

\begin{boxtheorem}
    If $\alpha$ is an infinite ordinal, then $\abs{\alpha \times \alpha} = \abs{\alpha}$. That is, there is a bijection between $\alpha \times \alpha$ and $\alpha$.
\end{boxtheorem}
\begin{proof}
    First, observe that if $\alpha$ is countably infinite, then the result holds, because we know that the Cartesian product of countably infinite sets is countably infinite. We focus our attention on the case where $\alpha$ is uncountably infinite.

    \sorry
\end{proof}

\begin{boxcorollary}
    If $\parenth{A; \le}$ is an infinite well-ordered set, then $\abs{A \times A} = \abs{A}$.
\end{boxcorollary}
\begin{proof}
    By \sorry, we know that there is an ordinal $\alpha$ with $\parenth{A; \le} \simeq \parenth{\alpha; \in}$. Therefore,
    \begin{align*}
        \abs{A \times A} = \abs{\alpha \times \alpha} = \abs{\alpha} = \abs{A}
    \end{align*}
\end{proof}

\subsection{Transfinite Recursion}

In the natural numbers, there is more than one use of Peano's Fifth Axiom: it is used not only to prove results by induction, but it is also used to construct things inductively. In this subsection, we explore the notion of transfinite recursion, which provides us with a similar framework that works over ordinals instead of just natural numbers.

The objective of transfinite recursion is to allow us to construct, for ordinals $\alpha$, sets $G(\alpha$ that are obtained from $G(\beta)$ (for $\beta < \alpha$) by applying some operation $F$ (cf. \Cref{Ch3:Def:Op_on_Sets}). In other words, we want to find an operation on sets $F$ that allows us to construct a family of sets $G(\cdot)$, indexed by ordinals, such that $F$, when applied to the sets $G(0), G(1), \ldots, G(\beta), \ldots$ indexed by ordinals that are less than $\alpha$, gives $G(\alpha)$.



\begin{boxtheorem}[The Principle of Transfinite Recursion]\label{Ch3:Thm:Transfinite_Recursion}
    Let $F$ be an operation on sets. Then, there is an operation $G$ such that for all ordinals $\alpha$, we have
    \begin{align*}
        G(\alpha) = F\of{G \tfr \alpha}
    \end{align*}
    Furthermore, $G$ is unique on ordinals, in the sense that if $G'$ is another operation such that
    \begin{align*}
        G'(\alpha) = F\of{G' \tfr \alpha}
    \end{align*}
    for all ordinals $\alpha$, then for all ordinals $\alpha$,
    \begin{align*}
        G(\alpha) = G'(\alpha)
    \end{align*}
\end{boxtheorem}

The proof is quite technical, and was omitted from lectures. It can be found in \cite[Appendix A.4]{LecNotes2025}.

Note that the operation $G$ given in the statement of \Cref{Ch3:Thm:Transfinite_Recursion} is \underline{not unique in an absolute sense}: we have no control over how $G$ behaves on sets that are not ordinals. \Cref{Ch3:Thm:Transfinite_Recursion} only tells us that $G$ is unique \underline{on ordinals}.

An interesting application of the principle of transfinite recursion is the following proof of the Lindenbaum Lemma (cf. \Cref{Ch2:Prop:Lindenbaum_FO}) for languages whose alphabet is well-ordered.

\begin{boxcorollary}[Lindenbaum Lemma for Well-Ordered Languages]
    Let $\L$ be a first-order language whose alphabet of symbols is well-ordered (cf. \sorry). If $\Sigma$ is a set of closed $\KL$-formulae, then there is a consistent set $\Sigma^*$ of closed $\KL$-formulae such that $\Sigma \subseteq \Sigma^*$.
\end{boxcorollary}
\begin{proof}
    \sorry
\end{proof}

